---
title: "Week 11: Causal Inference"
subtitle: "Multilevel Causal Inference"
execute:
  echo: false
params:
  slides: "11_causal_inference"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Define **causal effect** from a causal inference framework
- Describe what a confounder is using a **directed acyclic graph (DAG)**
- Explain how **randomized experiments** control for confounders
- Explain when and how **statistical adjustment** can potentially remove confounding
- Explain how including **cluster means** can remove confounders at level 2

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * Rhoads & Li (2022) (to be shared on Slack)
    * [Feller & Gelman (2015)](http://www.stat.columbia.edu/~gelman/research/published/HierarchicalCausal.pdf)
3. Attend the Tuesday session on last week's exercise (cross-lagged models)
4. Attend the Thursday session and participate in the class exercise
5. Complete Homework 9

# Lecture

## Slides

```{r}
xaringanExtra::embed_xaringan(file.path("/slides", paste0(params$slides, ".html")))
```

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Overview

{{< video https://youtu.be/1fixYnYj6Vk >}}

## Causal Inference Frameworks

{{< video https://youtu.be/JEtIswjQwm4 >}}

```{r}
#| results: asis
practice_question(
    "In a study of the association between chocolate consumption and number of Nobel laureates, what would be considered a causal effect?",
    answer = "The change in the number of Nobel laureates if everyone in the US ate more chocolate while keeping everything else constant",
    options = c("The difference in the number of Nobel laureates between countries with high and low per capita chocolate consumptions",
                "The difference in the number of Nobel laureates between people, in the same age group, with high and low average chocolate consumptions",
                "The change in the number of Nobel laureates if everyone in the US ate more chocolate while keeping everything else constant"),
    type = "radio"
)
```

```{r}
#| results: asis
practice_question(
    "A potential outcome for a person i is",
    answer = "The outcome value the same person would have obtained if they were in a particular treatment condition",
    options = c("The outcome value the same person would have obtained if they were in a particular treatment condition",
                "The average outcome value of everyone similar to person i",
                "What person i could have obtained if the treatment is made stronger"),
    type = "radio"
)
```

## Confounding

{{< video https://youtu.be/jrShzRJ10dM >}}

::: {.callout-note icon=false}

## Check your learning

```{r, message = FALSE}
library(dagitty)
library(ggdag)
dag1 <- dagitty(
    "dag{
      X -> Y; U2 -> X; U2 -> Y
      X -> U1; U1 -> Y
      U3 -> X; U4 -> Y
    }"
)
coordinates(dag1) <- list(x = c(X = 0, U1 = 0, U2 = 0.5, U3 = 1, U4 = 1.5, Y = 2),
                          y = c(X = 0, U1 = 1, U2 = 1, U3 = 1, U4 = 1, Y = 0))
# Plot
ggdag(dag1) + theme_dag()
```

In the graph above, which variable is a confounder?

```{r}
#| results: asis
check_question("U2", 
               options = c("U1", "U2", "U3", "U4"), 
               type = "radio", right = right_ans, wrong = wrong_ans, 
               alignment = TRUE)
```

:::

{{< video https://youtu.be/J47ppYZ-Tdc >}}

```{r}
#| results: asis
practice_question(
    "In an experiment, why do we only get an average treatment effect (ATE), instead of causal effects of individuals?",
    answer = "Because we only observe one potential outcome for each individual",
    options = c("Because the average is more accurate than individual scores",
                "Because we only observe one potential outcome for each individual",
                "Because the counterfactual score of each person in the treatment group is the same as the mean of the control group"),
    type = "radio"
)
```

## Statistical Adjustment

{{< video https://youtu.be/HNvCedHr0nU >}}

::: {.callout-note icon=false}

## Check your learning

```{r}
set.seed(2134)
x <- runif(100)
e <- rnorm(100)
e <- (e - mean(e)) / sd(e) * 0.5
m <- dplyr::ntile(x, 5)
y <- 0.5 + 0.5 * x + m + e
library(ggplot2)
tibble::tibble(y, x, m) |>
    ggplot(aes(x = x, y = y, col = factor(m))) +
    geom_point() +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(col = "m")
```

Which of the following is closest to the effect of x on y, adjusting for m?

```{r}
#| results: asis
check_question("0", 
               options = c("0", "0.5", "-1"), 
               type = "radio", right = right_ans, wrong = wrong_ans, 
               alignment = TRUE)
```

:::

{{< video https://youtu.be/-0HJwm1Mww8 >}}

::: {.callout-note icon=false}

## Check your learning

```{r, message = FALSE}
library(dagitty)
library(ggdag)
dag1 <- dagitty(
    "dag{
      X -> Y; U2 -> X; U2 -> Y
      X -> U1; U1 -> Y
      U3 -> X; U4 -> Y
    }"
)
coordinates(dag1) <- list(x = c(X = 0, U1 = 0, U2 = 0.5, U3 = 1, U4 = 1.5, Y = 2),
                          y = c(X = 0, U1 = 1, U2 = 1, U3 = 1, U4 = 1, Y = 0))
# Plot
ggdag(dag1) + theme_dag()
```

In the graph above, which variable is a mediator?

```{r}
#| results: asis
check_question("U1", 
               options = c("U1", "U2", "U3", "U4"), 
               type = "radio", right = right_ans, wrong = wrong_ans, 
               alignment = TRUE)
```

:::

## Multilevel Causal Inference

### Berkeley admission example

{{< video https://youtu.be/5hGh8ZVMRsI >}}

{{< video https://youtu.be/7s4yuSARTuQ >}}

### The role of cluster means

{{< video https://youtu.be/Jc7xFgRzhEk >}}

```{r}
#| results: asis
practice_question(
    "From a causal inference perspective, why is including the cluster mean beneficial?",
    answer = "Because confounding variables at the cluster level have effects on the cluster mean, which is held constant by including in the model",
    options = c("Because confounding variables at the cluster level have effects on the cluster mean, which is held constant by including in the model",
                "Because the cluster mean is a mediator variable",
                "Because the between-cluster and within-cluster effects can be decomposed"),
    type = "radio"
)
```

:::