---
title: "Week 2"
subtitle: "Review of Regression"
execute:
  echo: false
params:
  slides: "02_review_regression"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Refresher Learning Objectives

By the end of this module, you will be able to

- Identify alternative names for multilevel modeling (MLM)
- Describe the types of data MLM can handle
- Knit a simple R Markdown file
- Review concepts in multiple regression, including model equations, interactions, and interpretations

## Slides

```{r}
xaringanExtra::embed_xaringan(file.path("/slides", paste0(params$slides, ".html")))
```

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Statistical Model

```{r}
add_youtube("axg_AshO7qM")
```

```{r}
#| results: asis
practice_question(
    "In the example in the video, why do we need a random component?",
    answer = "Because people who spend the same amount of time studying do not always have the same mastery level",
    options = c("Because there is only one predictor",
                "Because the relationship between the variables is linear",
                "Because people who spend the same amount of time studying do not always have the same mastery level"),
    type = "radio"
)
```

## Import Data

```{r}
add_youtube("NgjdsKxLMbI")
```

Check your learning: What is the coding for the `sex` variable?

```{r}
#| results: asis
practice_question(
    "What is the coding for the `sex` variable?",
    answer = "0 = male, 1 = female",
    options = c("1 = male, 2 = female",
                "2 = male, 1 = female",
                "0 = female, 1 = male",
                "0 = male, 1 = female"),
    type = "radio"
)
```

```{r}
add_youtube("UfFhGF3mJGk")
```

::: {.callout-note icon=false}

Take a pause and look at the scatterplot matrix. Ask yourself the following:

- How does the distribution of `salary` look?
- Are there more males or females in the data?
- How would you describe the relationship between number of publications and salary?

:::

## Linear Regression

```{r, echo=FALSE}
add_youtube("JcmmFSruAZA")
```

### Sample regression line

```{r, echo=FALSE}
add_youtube("GIRtlYGVOXk")
```

```{r}
#| results: asis
practice_question(
    "How would you translate the regression line $y = \\beta_0 + \\beta_1 \\text{predictor1}$ into R?",
    answer = "y ~ predictor1",
    options = c("y ~ 0 + predictor1",
                "predictor1 ~ 1 + y",
                "y ~ predictor1"),
    type = "radio"
)
```

### Centering

```{r}
add_youtube("s19aoHq53zw")
```

```{r}
#| results: asis
practice_question(
    "The mean of the `pub` variable is 18.2. If we call the mean-centered version of it as `pub_c`, what should be the value of `pub_c` for someone with 10 publications?",
    answer = " -8.2",
    options = c(" 8.2", " 1.8", " -8.2", " 1.82"),
    type = "radio",
    alignment = TRUE
)
```

### Categorical Predictor

```{r}
add_youtube("k2nNMfJ2Xn0")
```

```{r echo=FALSE, results="asis"}
```{r}
#| results: asis
practice_question(
    "In a regression analysis, assume that there is a binary predictor that indicates whether a school is public (coded as 0) or private (coded as 1). If the coefficient for that predictor is 1.5, which category has a higher predicted score?",
    answer = "private",
    options = c("private", "public"),
    type = "radio",
    alignment = TRUE
)
```

## Multiple Regression

```{r}
add_youtube("O2ob6wC91w0")
```

::: {.callout-note icon=false}

## Think more

The coefficient of `pub_c` becomes smaller after adding `time` into the equation. Why do you think that is the case?

:::

### Interaction

```{r}
add_youtube("cqGZD8s5WvU")
```

```{r}
add_youtube("vK-2xMftORM")
```

::: {.callout-note icon=false}

## Practice yourself

From the interaction model, obtain the regression line when `pub` = 50.

:::

:::