---
title: "Week 1: Overview"
subtitle: "Overview of Multilevel Models"
execute:
  echo: false
params:
  slides: "01_intro"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

Before starting the materials for week 1, make sure you have reviewed the [syllabus](/syllabus.qmd).

# Week Learning Objectives

By the end of this module, you will be able to

- Identify alternative names for multilevel modeling (MLM)
- Describe the types of data MLM can handle
- Knit a simple R Markdown file

## Task Lists

1. Attend the Tuesday session for course introduction and Q&A
2. Install/Update R and RStudio on your computer
3. Review the resources (lecture videos and slides)
4. Complete the assigned readings
    * SB ch 1, [2][3]
    * [R Markdown Intro][1]
    * [rmarkdown cheatsheet][2]
    * - [Gelman et al. ch 1.1, 1.2, 1.4][4]
5. Introduce yourself on the #introduction Slack channel (as part of HW 1)
6. Attend the Thursday session and participate in the class exercise
7. Complete Homework 1

[1]: https://rmarkdown.rstudio.com/lesson-1.html
[2]: https://raw.githubusercontent.com/rstudio/cheatsheets/main/rmarkdown.pdf
[3]: https://us.sagepub.com/sites/default/files/upm-assets/44075_book_item_44075.pdf
[4]: https://statmodeling.stat.columbia.edu/wp-content/uploads/2020/07/raos_overview.pdf

# Lecture

## Slides

```{r}
xaringanExtra::embed_xaringan(file.path("/slides", paste0(params$slides, ".html")))
```

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Multilevel Data Structure

```{r}
add_youtube("9TtyDmczGGE")
```

```{r}
#| results: asis
practice_question(
    "How would you describe the data structure in the video?",
    answer = "hierarchical",
    options = c("observational", "educational", "hierarchical"),
    type = "radio",
    alignment = TRUE
)
```

## Name and History

```{r}
add_youtube("9Ax8W6VPgmk")
```

```{r}
#| results: asis
practice_question(
    "What is another name for multilevel modeling?",
    answer = c("mixed-effect modeling"),
    options = c("mixed-effect modeling", "hierarchical regression",
                "fixed-effect modeling"),
    type = "radio"
)
```

## Usage of MLM

```{r}
add_youtube("4fl2chIkDy4?end=340")
```

::: {.callout-tip appearance="simple"}

Another example that relates to ecological fallacies is the ["happiness paradox"](https://www.pnas.org/doi/10.1073/pnas.1015962107) in economics, which says that income is related to happiness in cross-sectional analysis (i.e., between-person level) but not in a longitudinal/time series analysis (i.e., within-person level).

:::

```{r}
#| results: asis
practice_question(
    "In the data structure where there are multiple measurements for the same person, which level is level 1?",
    answer = c("measurement"),
    options = c("measurement", "person"),
    type = "radio",
    alignment = TRUE
)
```

```{r}
add_youtube("cb0YL4LqUN4")
```

:::