---
title: "Week 6"
subtitle: "Model Diagnostics and Results Reporting"
execute:
  echo: false
params:
  slides: "06_model_diagnostics"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Describe the major **assumptions** in basic multilevel models
- Conduct analyses to decide whether **cluster means** and **random slopes** should be included
- Use graphical tools to diagnose assumptions of **linearity**, **homoscedasticity** (equal variance), and **normality**
- Solve some basic **convergence issues**
- **Report** results of a multilevel analysis based on established guidelines

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * Snijders & Bosker ch 10
    * Meteyard & Davies (2020; to be shared on Slack)
    * [McCoach (2019 chapter)](https://www-taylorfrancis-com.libproxy2.usc.edu/books/e/9781315755649/chapters/10.4324/9781315755649-22) (USC SSO required)
3. Attend the Tuesday session to learn some R skills and review last week's exercise
4. Attend the Thursday session and participate in the class exercise
4. Complete Homework 5

# Lecture

## Slides

```{r}
xaringanExtra::embed_xaringan(file.path("/slides", paste0(params$slides, ".html")))
```

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Model Diagnostics

```{r}
add_youtube("YcwLBSTY8aw")
```

```{r}
#| results: asis
practice_question(
    "Homoscedasticity means",
    answer = "Equal variance of errors",
    options = c("Equal variance of errors",
                "Each cluster has the same mean",
                "The model resembles the data"),
    type = "radio"
)
```

## Assumptions of a Multilevel Model

```{r}
add_youtube("dhh5twY9pwE")
```

Note: $\mathrm{E}(Y)$ can also be written as $\hat Y$, the predicted value of $Y$ based on the predictor values.

The linear model is also flexible as it can allow predictors that are curvillinear terms, such as $Y = b_0 + b_1 X_1 + b_2 X_1^2$, or $Y = b_0 + b_1 \log(X_1)$, or more generally
$$Y = b_0 + \sum_{i}^p b_i f(x_1, x_2, \ldots)$$
The "linear" part in a linear model actually means that $Y$ is a linear function of the coefficients $b_1, b_2, \ldots$.

The second functional form in the slide, however, is a truly nonlinear function.

```{r}
#| results: asis
practice_question(
    "Which of the following is **NOT** a linear model?",
    answer = "Y = b0 * exp(b1 * X1 + b2 * X2)",
    options = c("Y = X1 + X2",
                "Y = b0 + b1 * X1 + b2 * X2 + b3 * X1 * X2",
                "Y = b0 * exp(b1 * X1 + b2 * X2)"),
    type = "radio"
)
```

```{r}
#| results: asis
practice_question(
    "What is implied when the model specifies that the variance of $u_{0j}$ is $\\tau^2_0$?",
    answer = "The intercept variance is constant across different clusters",
    options = c("The intercept variance is constant across different clusters",
                "The slope is different across clusters",
                "The intercept variance is different for different types of schools"),
    type = "radio"
)
```

### Remember the "LINES" 

```{r}
add_youtube("lh_5-2Cri-4")
```

```{r}
#| results: asis
practice_question(
    "What does \"I\" in \"LINES\" stand for?",
    answer = "Independence of errors",
    options = c("Increasing trend of outcome",
                "Independence of errors",
                "Independence of predictors"),
    type = "radio")
```

```{r}
add_youtube("TXwXrqbK8vE")
```

```{r}
#| results: asis
practice_question(
    "What is shown in a marginal model plot?",
    answer = "Whether the model-implied association is similar to the model-free (i.e., marginal) association",
    options = c("Whether the model shows a linear association between a predictor and the outcome",
                "Whether the model-implied association is similar to the model-free (i.e., marginal) association",
                "Whether the margin of error in the model is small enough"),
    type = "radio"
)
```

```{r}
add_youtube("2kf5g_WzGhk")
```

```{r}
add_youtube("4GKjDRYcWyM")
```

::: {.callout-note icon=false}

## Check your learning

Which assumption(s) are likely violated in the following plot?

```{r practice-assumption, message = FALSE}
library(tidyverse)
library(lme4)
library(broom.mixed)
set.seed(10)
u0 <- abs(rnorm(20))
x <- runif(200)
e <- rnorm(200, sd = pmin(1 / sqrt(x), 5))
id <- rep(seq_len(20), each = 10)
y <- 0.5 + 2 * log(x) + u0[id] + e
m1 <- lmer(y ~ x + (1 | id))
augment(m1) %>%
    mutate(.ols_resid = .fixed - y) %>%
    ggplot(aes(x = x, y = .ols_resid)) +
    geom_point() +
    geom_smooth(se = FALSE)
```

```{r}
#| results: asis
check_question(
    "linearity and equal variance of errors",
    options = c("linearity",
                "equal variance of errors",
                "normality",
                "linearity and equal variance of errors"),
    type = "radio", right = right_ans, wrong = wrong_ans
)
```

:::

## Other Model Issues

::: {.callout-note icon=false}

## Outliers/influential observations

- Check coding error
- Don't drop outliers unless you adjust the standard errors accordingly, or use robust models (see an example in the [R code](/codes/rcode6.qmd#bonus-bayesian-robust-mlm))

:::

::: {.callout-note icon=false}

## Reliability (e.g., $\alpha$ coefficient)

- Reliability may be high at one level but low at another level
- See Lai (2021, doi: 10.1037/met0000287) for level-specific reliability
    * You can use the `multilevel_alpha()` function from https://github.com/marklhc/mcfa_reliability_supp/blob/master/multilevel_alpha.R
    * The procedure was recently implemented in the `semTools::compRelSEM()` function, thanks to Dr. Terrence D. Jorgensen.

:::

## Handling Convergence Issues

```{r}
add_youtube("9wtTLahkGFE")
```

See [this section in the R code](/codes/rcode6.qmd#convergence)

## Reporting Results

```{r}
add_youtube("12tK6d468u0")
```

:::