---
title: "Week 8: Cross-Classified Models"
subtitle: |
  Multilevel Models for Experimental Data
execute:
  echo: false
params:
  slides: "08_mlm_for_experimental_data"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Identify the correct levels with experimental studies
- Describe designs with crossed random levels
- Assign variables to appropriate levels, and tell which variables can have random slopes at which levels
- Compute a version of effect size ($R^2$) for experimental data

::: {.callout-note}

In the video, the slides use $d$ for effect size, but we will use $R^2$ in this class.

:::

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * Snijders & Bosker 13.1
    * [Hoffman & Rovine (2007)](https://link.springer.com/content/pdf/10.3758/BF03192848.pdf)
3. Attend the **Tuesday** session and participate in the class exercise
5. Complete the project prospectus
4. Complete Homework 7 (due in two weeks)
6. Additional resources for learning MLM for experimental designs
    * [This paper](https://www.annualreviews.org/doi/abs/10.1146/annurev-psych-122414-033702) by Judd et al. 

# Lecture

## Slides

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Multilevel Experiments

```{r}
add_youtube("9btZ_FrDGus")
```

```{r}
#| results: asis
practice_question(
    "In a research study, 10 hospitals are randomly assigned to a treatment condition to adopt a new drug, whereas the other 10 hospitals use the conventional method. What is the design of this study?",
    answer = "A cluster-randomized trial",
    options = c("A cluster-randomized trial",
                "A multisite trial",
                "A within-subjects design"),
    type = "radio"
)
```

## Example

```{r}
add_youtube("4baFBogxmAs")
```

```{r}
#| results: asis
practice_question(
    "In the data set, how many observations are there at level 1?",
    answer = "7,803",
    options = c("7,803",
                "153",
                "51"),
    type = "radio"
)
```

## Long vs. Wide data set

```{r}
add_youtube("NsRnhNz3-10")
```

::: {.callout-note icon=false}

## Check your learning

In the following data, `hvltt` to `hvltt4` are the test scores of a verbal learning test across four time points. Is this a long or a wide data set?

```{r}
#| message: false
library(tidyverse)
active_csv <- here::here("data_files", "active.csv")
if (!file.exists(active_csv)) {
    download.file(
        "https://books.psychstat.org/rdata/data/active.csv",
        active_csv
    )
}
read.csv(active_csv, header = TRUE) %>%
    select(id, hvltt:hvltt4) %>%
    rmarkdown::paged_table()
```


```{r}
#| results: asis
check_question("wide", 
               options = c("long", 
                           "wide"), 
               type = "radio", right = right_ans, wrong = wrong_ans, 
               alignment = TRUE)
```

:::

## Crossed Random Levels

```{r}
add_youtube("Vgv78TgDWgA?end=250")
```

```{r}
add_youtube("193WbUEpKuI")
```

```{r}
add_youtube("Vgv78TgDWgA?start=312")
```

::: {.callout-note icon=false}
## Think more

What is the data structure if there are 1,000 students from 100 schools and 30 neighborhoods, and each school has students from multiple neighborhoods?

:::

### Unconditional model

```{r}
add_youtube("zECzzA4H_uo")
```

:::{.callout-important appearance="simple"}

When there is > 1 observations per cell, one can (and **should**) estimate the interaction variance component using

```{r}
#| eval: false
#| echo: true
lg_rt ~ (1 | id) + (1 | Item) + (1 | id:Item)
```

:::

::: {.callout-tip icon=false}

## Practice yourself

Compute the design effects for the person level and for the item level. Do the design effects suggest the need for both levels? (That is, are both design effects > 1.1?)

Answer: See the computation in the [R code](rcode7.html#intraclass-correlations-and-design-effects)

:::

## Cross-Classified Random-Effect Model With Random Slopes

```{r}
add_youtube("dyDKHC0sugE")
```

```{r}
#| results: asis
practice_question(
    "If in the experiment, each person responds to each item 3 times, each time with a different `duration`. At what level(s) can `duration` have random slopes?",
    answer = "Both person and item", 
    options = c("Both person and item", 
                "Person only", 
                "Item only", 
                "Within-cell"), 
    type = "radio"
)
```

### Full Model

```{r}
add_youtube("bIrEtjB91gk")
```

## Effect Size

Please check out the slides and the examples in the [R code](/codes/rcode8.qmd#effect-size)

:::
