---
title: "Week 9: Longitudinal I"
subtitle: |
  Models for Longitudinal Data I
execute:
  echo: false
params:
  slides: "09_longitudinal_analysis_1"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Describe the similarities and differences between longitudinal data and cross-sectional clustered data
- Perform some basic **attrition analyses**
- Specify and run **growth curve analysis**
- Analyze models with **time-invariant covariates** (i.e., lv-2 predictors) and interpret the results

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * Snijders & Bosker ch 15 (you can skip 15.1.3 and 15.1.4)
3. Attend the Tuesday session to learn about `brms`
4. Attend the Thursday session and participate in the class exercise
4. Complete Homework 7 (on materials for Week 8)
5. Additional resources for learning MLM for longitudinal data analysis
    * [This excellent book by Hoffman (2014)](https://www-taylorfrancis-com.libproxy2.usc.edu/books/9781315744094) (USC SSO required)

# Lecture

## Slides

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Longitudinal Data Analysis

```{r, echo = FALSE}
add_youtube("bdDrcmktBnE")
```

```{r}
#| results: asis
practice_question(
    "In a research study, data were collected for a group of patients on symptoms of eating disorder on a weekly interval across 5 weeks. What type of data is this?",
    answer = "Panel data", 
    options = c("Panel data", 
                "Intensive longitudinal data", 
                "Cross-sectional data"), 
    type = "radio"
)
```

## Example data

```{r}
add_youtube("3PDA_ur1sOU")
```

```{r}
#| results: asis
practice_question(
    "In the data set, at what level is `homecog`, which is a measure of mother's cognitive stimulation at baseline?",
    answer = "Level 2", 
    options = c("Level 1", 
                "Level 2"), 
    type = "radio", 
    alignment = TRUE
)
```

### Basic attrition analysis

```{r}
add_youtube("b2dto5JwRQ8")
```

::: {.callout-note icon=false}

See the [R code section](/codes/rcode9.qmd#attrition-analyses).

:::

```{r}
#| results: asis
practice_question(
    "In the spaghetti plot, what does the average trend line mean?",
    answer = "That the sample average reading score increased across time", 
    options = c("That every child showed an increasing trend", 
                "That the sample average reading score increased across time", 
                "That there is no variability in individual trajectories"), 
    type = "radio"
)
```

## Growth Curve Modeling

```{r}
add_youtube("iT7kzf2gIUs")
```

```{r}
#| results: asis
practice_question(
    "In a growth model, what does it mean when $\\tau_1 = 0$?",
    ctitle = "Think more",
    answer = "That every participant has the same rate of change", 
    options = c("That on average, there is no change on the outcome across time", 
                "That every participant has the same rate of change", 
                "That every participant has the same outcome score"), 
    type = "radio"
)
```

### Linear growth

:::{.callout-important appearance="simple"}

In the videos, what is labelled as SD~post~ is the Bayesian analog of the standard error.

:::

```{r}
add_youtube("4xHz4OIPVh8")
```

```{r}
#| results: asis
practice_question(
    "What is the advantage of having time to start at 0?",
    answer = "The intercept corresponds to the predicted score at the first time point", 
    options = c("The intercept corresponds to the predicted score at the first time point", 
                "So that $\\tau_1$ will be the SD of the slopes"), 
    type = "radio"
)
```

### Piecewise linear growth

```{r}
add_youtube("p6TQdUzi1d0")
```

::: {.callout-note icon=false}

## Practice yourself

What should the coding of phase 1 and phase 2 be if the turning point is set at `time` = 2?

:::

```{r}
add_youtube("MbDcv05wOrU")
```

::: {.callout-note icon=false}

## Note

In this example, the turning point was chosen mostly based on the spaghetti plot and was arbitrary. For your research, you should justify your choice.

:::

```{r}
#| results: asis
practice_question(
    "If a piecewise growth model has an AIC of 23745, and a linear growth model has an AIC of 23650, which model should be preferred?",
    answer = "Linear",
    options = c("Piecewise", 
                "Linear"), 
    type = "radio",
    alignment = TRUE
)
```

## Time-Invariant Covariates 

```{r}
add_youtube("fI6XJnVQ3pg")
```

```{r}
#| results: asis
practice_question(
    "What does the coefficient for `phase1` mean when the model includes an interaction between `phase1` and `homecog9`?",
    ctitle = "Think more",
    answer = "The average slope from time 0 to time 1 for kids with homecog = 9",
    options = c("The average slope from time 0 to time 1",
                "The average slope from time 0 to time 1 for kids with homecog = 9",
                "The average slope from time 0 to time 1, holding constant homecog"),
    type = "radio"
)
```

## Varying Occasions

::: {.callout-note icon=false}

Instead of using `time` as the duration since a particular point in history (e.g., when the study started), one can use other ways of quantifying time, such as the duration since one is born (i.e., chronological age). See R code.

:::

```{r}
add_youtube("XsbxN2m_si4")
```

:::{.callout-important appearance="simple"}

In the video below, recorded in 2021, I used the R package `glmmTMB` for frequentist analyses. The results and interpretations using `brms` are similar.

:::

```{r}
add_youtube("PYyMVi292DM")
```

:::
