---
title: "Week 4: Lv-1 Predictor"
subtitle: "Effect Decomposition, Random Coefficient Model, and Cross-level Interactions"
execute:
  echo: false
params:
  slides: "04_adding_level-1_predictor"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Explain what the ecological fallacy is
- Use **cluster-mean/group-mean centering** to decompose the effect of a lv-1 predictor
- Define **contextual effects**
- Explain the concept of **random slopes**
- Analyze and interpret **cross-level interaction** effects

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * Snijders & Bosker ch 4.6, 5.1--5.3
3. Attend the Tuesday Q&A + review session (optional)
4. Attend the Thursday session and participate in the class exercise
5. Complete Homework 4
6. Now that you have learned the basics of MLM, start thinking about your project (Prospectus due around Oct 17)

# Lecture

## Slides

[PDF slides](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Overview

```{r}
add_youtube("0bIXkUTEC3U")
```

```{r}
#| results: asis
practice_question(
    "The Type I error inflation problem when using OLS regression for clustered data applies to",
    answer = "Level-2 predictors and most level-1 predictors", 
    options = c("Level-1 predictors", 
                "Level-2 predictors", 
                "Level-2 predictors and most level-1 predictors"),
    type = "radio"
)
```

## Ecological Fallacy

```{r}
add_youtube("4SnA5Du4E-c")
```

::: {.callout-note icon=false}

## Check your learning

In the ["bizarre" research finding](https://blogs.scientificamerican.com/the-curious-wavefunction/chocolate-consumption-and-nobel-prizes-a-bizarre-juxtaposition-if-there-ever-was-one/) that showed a correlation between chocolate consumption and number of Nobel prize winners at the country level, which of the following is reasonable to infer?

```{r}
#| results: asis
check_question(
    "That there are potential third variables that relate to both a country's chocolate consumption level and the number of Nobel prize winners", 
    options = c("That those who eat more chocolate are more likely to win a nobel prize", 
                "That nutritional contents in chocolate is related to cognitive functioning", 
                "That there are potential third variables that relate to both a country's chocolate consumption level and the number of Nobel prize winners"),
    type = "radio", right = right_ans, wrong = wrong_ans
)
```

:::

::: {.callout-note icon=false}

## Check your learning

Summarize the "Big-Fish-Little-Pond Effect" in terms of how a person's own academic performance and the overall performance of the person's school on academic self-concept.

:::

## Decomposing Effects

### Between/within effects

::: {.callout-note icon=false}

What I called "cluster-mean centering" is the same as "within-group centering" in Snijders & Bosker (2012)

:::

```{r}
add_youtube("6h1U6aZmohY")
```

```{r}
#| results: asis
practice_question(
    "Why do we need to separate a level-1 predictor into two variables in the model?",
    answer = "Because the slope of the level-1 predictor and that of its cluster means may be different", 
    options = c("Because the slope of the level-1 predictor and that of its cluster means may be different", 
                "Because there is an interaction between the predictor and the cluster variable", 
                "Because the intraclass correlation for any level-1 predictor will be zero"),
    type = "radio"
)
```

### Path diagram and equations

```{r}
add_youtube("ZhLOtzzxIRk")
```

::: {.callout-note icon=false}

## Think more

Based on the between-cluster level component in the path diagram and in the equations, `meanses` can predict

```{r}
#| results: asis
check_question(
    "Only the school means of math achievement", 
    options = c("Both the school-level math achievement and a student's relative standing within a school", 
                "Only a student's relative standing within a school", 
                "Only the school means of math achievement"),
    type = "radio", right = right_ans, wrong = wrong_ans
)
```

:::

```{r}
#| results: asis
practice_question(
    "Based on the results shown in the video, is the school-level slope or the student-level slope larger for the association between SES and math achievement?",
    answer = "School-level", 
    options = c("School-level", 
                "Student-level"),
    type = "radio"
)
```

### Interpret the between/within effects

```{r}
add_youtube("Pet7lt7Nb68")
```

::: {.callout-note icon=false}

## Try it yourself

Obtain the predicted `mathach` for Student B, and compare with Students A and C. 

:::

## Contextual Effects

```{r}
add_youtube("TTCE9Tuqj0Y")
```

Check your learning: The contextual effect is

```{r}
#| results: asis
practice_question(
    "The contextual effect is",
    answer = "The predicted difference in the outcome for every unit difference in the cluster mean predictor, holding constant the level-1 predictor", 
    options = c("The predicted difference in the outcome for every unit difference in the cluster mean predictor, holding constant the level-1 predictor", 
                "The predicted difference in the outcome for every unit difference in the level-1 predictor, within a given cluster", 
                "The between-level effect"),
    type = "radio"
)
```

## Random Slopes/Random Coefficients

### Developing intuition

```{r}
add_youtube("K6Q9wbsdgXw")
```

```{r}
#| results: asis
practice_question(
    "In a random-coefficient model, if there are $J$ cluster, there are",
    answer = "J intercepts, J slopes", 
    options = c("one intercept, two slopes", 
                "one intercept, J slopes", 
                "J intercepts, one slopes", 
                "J intercepts, J slopes"),
    type = "radio"
)
```

### Equations and path diagram

```{r}
add_youtube("_jXGrTbHft4")
```

::: {.callout-note icon=false}

## Check your learning

Which combination of $\tau_0$ and $\tau_1$ best describes the graph below?

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
ggplot(data = tibble(x = c(-1, 3), y = c(-1, 1)), 
       aes(x = x, y = y)) + 
    geom_blank() + 
    geom_abline(data = tibble(a = rnorm(30, sd = 0.1), 
                              b = rnorm(30, mean = -0.2, sd = 0.2)), 
                aes(intercept = a, slope = b))
```


```{r}
#| results: asis
check_question("tau_0 = 0.1, tau_1 = 0.2", 
               options = c("tau_0 = 0.1, tau_1 = 0.2", 
                           "tau_0 = 0, tau_1 = 0.2", 
                           "tau_0 = 0.1, tau_1 = 0", 
                           "tau_0 = 0, tau_1 = 0"), 
    type = "radio", right = right_ans, wrong = wrong_ans
)
```

:::

### Interpretations

```{r}
add_youtube("o6PDE4IEFfg")
```

```{r}
#| results: asis
practice_question(
    "In a random-slope model, if $\\gamma_{10}$ (the average slope) = 0.2, $\\tau^2_1 = 0.04$, what is the 68% plausible range for the slopes across clusters?",
    answer = "[0, 0.4]", 
    options = c("[0, 0.4]", 
                "[-0.48, 0.88]", 
                "[0.16, 0.24]", 
                "[0.12, 0.28]"),
    type = "radio"
)
```

## Cross-Level Interaction

<span style="color:red">In the video, there was a mistake in the path diagram, in that one of the circle should be $\beta_{1j}$, not $\beta_{0j}$</span>

```{r}
add_youtube("7RzVD02fZHw")
```

```{r}
#| results: asis
practice_question(
    "Conceptually, a cross-level interaction is the same as",
    answer = "Using a cluster-level predictor to predict variations in level-1 slopes", 
    options = c("An interaction between two level-1 predictors",
                "Using a cluster-level predictor to predict variations in level-1 slopes", 
                "Random slopes", 
                "Adding the main effect of a cluster-level predictor"),
    type = "radio"
)
```

:::
