---
title: "Week 12: Logistic Models"
subtitle: "Multilevel Logistic Models"
execute:
  echo: false
params:
  slides: "12_multilevel_logistic"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to

-   Describe the problems of using a regular multilevel model for a binary outcome variable
-   Write model equations for multilevel logistic regression
-   Estimate intraclass correlations for binary outcomes
-   Plot model predictions in probability unit

## Task Lists

1.  Review the resources (lecture videos and slides)
2.  Complete the assigned readings
    -   Snijders & Bosker ch 17.1--17.4
3.  Attend the Thursday session and participate in the class exercise
4.  Complete peer review for two of your peers' postings (due Monday, November 15, end of the day)
5.  Complete Homework 9 <!-- 6. (Optional) Read the [bonus material](#) on binomial logistic models -->

# Lecture

## Slides

[Note that in some of the videos below, the Bayesian analyses were used; however, for the class this year, we will stay with frequentist analyses. The results and interpretations are basically; just note some differences in the terminology.]{style="color:red"}

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

::: {.panel-tabset}

## Logistic MLM

```{r}
add_youtube("amg_pAGFS-A")
```

```{r}
#| results: asis
practice_question(
  "Which of the following is a binary variable?",
  answer = "Whether a participant gets hospitalized or not.", 
  options = c("The number of symptoms a person show per day.", 
              "The time from the start of a study to when a participant gets hospitalized.", 
               "Whether a participant gets hospitalized or not."), 
  type = "radio"
)
```

### Example Data

```{r}
add_youtube("LSDPVPUQpUc")
```

```{r}
#| results: asis
practice_question(
  "Which of the following is **not** a reason to use a logistic model?",
  answer = "Dichotomizing an outcome can make interpretation easier.", 
  options = c("Dichotomizing an outcome can make interpretation easier.", 
              "It uses a distribution suitable for a binary outcome.", 
              "It accommodates heterogeneous variance."), 
  type = "radio"
)
```

## Elements of Logistic MLM

```{r}
add_youtube("ZWDSTPnahVs")
```

```{r}
#| results: asis
practice_question(
  "A logistic model assumes that the outcome follows",
  answer = "a Bernoulli distribution.", 
  options = c("a normal distribution.", 
              "a Student's t distribution.", 
              "a Bernoulli distribution."), 
  type = "radio"
)
```

```{r}
add_youtube("fIZZSmVdiIs")
```

```{r}
#| results: asis
practice_question(
  "A log odds of 0.5 corresponds to a probability of",
  answer = "0.62.", 
  options = c("0.12.", 
              "0.5", 
              "0.62"), 
  type = "radio"
)
```

You are not alone if you have trouble understanding "odds" and "log odds". This video might give you a better idea: <https://www.youtube.com/watch?v=ARfXDSkQf1Y>

## Unconditional Model

### Model equations

```{r}
add_youtube("oi79Zl-IQf8")
```

```{r}
#| results: asis
practice_question(
  "In a logistic model predicting whether a person reported a daily stressor, the coefficient of age was -0.5. The interpretation is that",
  answer = "every unit increase in age is associated with a 0.5 decrease in the log odds of reporting a daily stressor.", 
  options = c("every unit increase in age is associated with a 0.5 unit decrease of a daily stressor.", 
              "every unit increase in age is associated with a 0.5 percentage point decrease of reporting a daily stressor.", 
              "every unit increase in age is associated with a 0.5 decrease in the log odds of reporting a daily stressor."), 
  type = "radio"
)
```

## Adding Predictors

```{r}
add_youtube("P95ruzXQAiU")
```

## Generalized Linear Mixed Model (GLMM)

```{r}
add_youtube("Tq1XaZO_8Bc")
```

```{r}
#| results: asis
practice_question(
  "A researcher wants to compare the proportion of minority hires in the past year across departments and schools. The total number of hires for each department is known. Which model is the most appropriate?",
  answer = "Binomial logistic", 
  options = c("Binomial logistic", 
              "Poisson", 
              "Exponential"), 
  type = "radio"
)
```

:::