---
title: "Week 1"
subtitle: "Overview of Multilevel Models"
execute:
  echo: false
---

```{r}
#| include: false
source("../helper.R")
```

Before you start the materials for week 1, make sure you have reviewed the [syllabus](../syllabus.qmd). 

# Week Learning Objectives

By the end of this module, you will be able to 

- Identify alternative names for multilevel modeling (MLM)
- Describe the types of data MLM can handle
- Knit a simple R Markdown file

## Task Lists

1. Attend the Tuesday session for course introduction and Q&A
2. Install/Update R and RStudio on your computer
3. Review the resources (lecture videos and slides)
4. Complete the assigned readings
    * Snijders & Bosker chapters 1 and [2][4] (they're relatively short)
    * [R Markdown Intro][1]
    * [Markdown Quick Reference][2]
5. Introduce yourself on the #introduction Slack channel (as part of HW 1)
6. Attend the Thursday session and participate in the class exercise
7. Complete Homework 1

[1]: https://vimeo.com/178485416
[2]: https://web.mit.edu/r/current/RStudio/resources/markdown_help.html
[4]: https://us.sagepub.com/sites/default/files/upm-assets/44075_book_item_44075.pdf

# Lecture

## Slides

## Multilevel Data Structure

```{r}
add_youtube("ufvCUaUkYOA")
```

```{r}
#| results: asis
practice_question(
    "How would you describe the data structure in the video?",
    answer = "hierarchical",
    options = c("observational", "educational", "hierarchical"),
    type = "radio",
    alignment = TRUE
)
```

## Name and History

```{r}
add_youtube("9Ax8W6VPgmk")
```

```{r}
#| results: asis
practice_question(
    "What is another name for multilevel modeling?",
    answer = c("mixed-effect modeling"),  
    options = c("mixed-effect modeling", "hierarchical regression", 
                "fixed-effect modeling"), 
    type = "radio"
)
```

## Usage of MLM

```{r}
add_youtube("4fl2chIkDy4")
```

::: {.callout-tip appearance="simple"}

Another example that relates to ecological fallacies is the ["happiness paradox"](https://esrc.ukri.org/about-us/50-years-of-esrc/50-achievements/the-easterlin-paradox/) in economics, which says that income is related to happiness in a cross-sectional analysis (i.e., between-person level) but not in a longitudinal/time series analysis (i.e., within-person level).

:::

```{r}
#| results: asis
practice_question(
    "In the data structure where there are multiple measurements for the same person, which level is level 1?",
    answer = c("measurement"),
    options = c("measurement", "person"),
    type = "radio",
    alignment = TRUE
)
```
