---
title: "Week 10: Longitudinal II"
subtitle: |
  Models for Longitudinal Data II
execute:
  echo: false
params:
  slides: "10_longitudinal_analysis_2"
---

```{r}
#| include: false
source("../helper.R")
source("../render_slides.R", local = TRUE)
```

# Week Learning Objectives

By the end of this module, you will be able to 

- Specify models with alternative error covariance structures
- Describe the difference between analyzing trends vs. analyzing **fluctuations** with longitudinal data
- Run analyses with **time-varying predictors** (i.e., level-1 predictors)
- Interpret and plot results

## Task Lists

1. Review the resources (lecture videos and slides)
2. Complete the assigned readings
    * [Hoffman (2014) ch 4.1](https://www-taylorfrancis-com.libproxy1.usc.edu/books/9781315744094/chapters/10.4324/9781315744094-12) (USC SSO required)
    * [Hoffman (2014) ch 8](https://www-taylorfrancis-com.libproxy1.usc.edu/books/9781315744094/chapters/10.4324/9781315744094-17) (USC SSO required)
3. Attend the Tuesday session to learn about `brms`
4. Attend the Thursday session and participate in the class exercise
5. Complete Homework 8
6. (Optional) Read the bonus R code on the generalized estimating equations (GEE) method
    * Additional reference: <https://journals.sagepub.com/doi/abs/10.3102/10769986211017480>

# Lecture

## Slides

[PDF version](`r file.path("/slides", paste0(params$slides, ".pdf"))`){target="_blank"}

:::{.callout-important appearance="simple"}

In the videos for this week, you will see that I used the R package `glmmTMB` for frequentist analyses for fitting models with autoregressive covariance structures. These are useful for getting quick results, but they may sometimes run into convergence issues. Using `brms` is generally more stable. 

:::

::: {.panel-tabset}

## Longitudinal Data Analysis II

```{r}
add_youtube("k5WyosL97B0")
```

## Temporal Covariance/Correlation

```{r}
add_youtube("pzsL1y5XsQ0")
```

```{r}
#| results: asis
practice_question(
    "Assume that the temporal correlation decreases with a longer time gap. A researcher collects data at baseline (Time 1), 3-month follow-up (Time 2), and then 5-month follow-up (Time 3). Which correlation should be strongest?",
    answer = "Between Time 2 and Time 3",
    options = c("Between Time 1 and Time 2",
                "Between Time 2 and Time 3",
                "Between Time 1 and Time 3"),
    type = "radio"
)
```

### Covariance Structure in MLM

OLS and RI-MLM/RM-ANOVA

```{r}
add_youtube("UKxumumURd0")
```

```{r}
#| results: asis
practice_question(
    "The random-intercept model/repeated-measures ANOVA assumes a specific temporal covariance structure. What is that structure called?",
    answer = "Compound Symmetry",
    options = c("Independence",
                "Autoregressive",
                "Autoregressive and Moving Average",
                "Compound Symmetry"),
    type = "radio"
)
```

Random Slopes

```{r}
add_youtube("UiBMcbNGbHk")
```

Autoregressive(1) error structure

```{r}
add_youtube("Z5ZghnvZkYI")
```

```{r}
#| results: asis
practice_question(
    "In an AR(1) covariance structure, what is the implied correlation between Time 2 and Time 4, if $\\rho = .4$?",
    answer = ".16",
    options = c(".4",
                ".2",
                ".16",
                ".064"),
    type = "radio", 
    alignment = TRUE
)
```

## Analyzing Dynamics

```{r}
add_youtube("xW_84TYK85g?start=105")
```

```{r}
add_youtube("F-NNPPAuPWE")
```

```{r}
#| results: asis
practice_question(
    "When analyzing a conversation between a couple, a researcher is interested in whether a person follows up the partner's complaints with positive or negative behaviors. Is this an example of studying trends or fluctuations?",
    answer = "Fluctuations", 
    options = c("Trends", 
                "Fluctuations"), 
    type = "radio", 
    alignment = TRUE
)
```

## Examples

### Model 1

```{r}
add_youtube("mDdyowlEUic")
```

```{r}
#| results: asis
practice_question(
    "In the model discussed in the video, what is the interpretation of the contextual effect of `mood1`?",
    answer = "The predicted difference in symptoms between two individuals with 1 unit difference in average negative mood on a day when both individuals have the same daily mood level", 
    options = c("The predicted difference in average symptoms between two individuals with 1 unit difference in average negative mood", 
                "The predicted difference in symptoms for an individual across two days with 1 unit difference in mood", 
                "The predicted difference in symptoms between two individuals with 1 unit difference in average negative mood on a day when both individuals have the same daily mood level"), 
    type = "radio"
)
```

### Model 2

```{r}
add_youtube("bbNaNi2hUXQ")
```

::: {.callout-note icon=false}

For the coefficients of `stressor` and `stressor_pm` in the above model, the coefficients are ones adjusting for the other predictors in the model (e.g., `mood1_pm`, `mood1_pmc`, `women`, and their interactions). 

:::

:::