---
title: "Multilevel Causal Inference"
subtitle: "PSYC 575"
author: "Mark Lai"
institute: "University of Southern California"
date: "2022/10/28 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "my-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  for_class: false
---

```{r, child = "slide_settings.RMarkdown"}
```

# Week Learning Objectives

- Define **causal effect** from a causal inference framework

- Describe what a confounder is using a **directed acyclic graph (DAG)**

- Explain how **randomized experiments** control for confounders

- Explain when and how **statistical adjustment** can potentially remove confounding

- Explain how including **cluster means** can remove confounders at level 2

---

# Reading

### Rhoads & Li (2022) Chapter: Causal Inference in Multilevel Settings

### Feller & Gelman (2015). Hierarchical Models for Causal Effects.

---

# Causal Inference

.center[

### When and how can we determine the causal effect of $X$ on $Y$?

]

--

E.g., Sector on Achievement coefficient with HSB data

```{r import_sav, message = FALSE}
library(dplyr)
# Read in the data (pay attention to the directory)
hsball <- haven::read_sav(here::here("data_files", "hsball.sav")) |>
    mutate(sector = factor(sector))
hsb_po <- hsball |>
    group_by(sector) |>
    slice(1:6) |>
    ungroup()
```

```{r, output.lines = c(2:4, 17:20)}
library(lme4)
lmer(mathach ~ sector + (1 | id), data = hsball) |>
    summary()
```

- The predicted difference in achievement between students in Catholic (`sector` = 1) vs. public schools (`sector` = 0)

--

- $\hat Y \mid X = 1 - \hat Y \mid X = 0$

---

# Causal Effect

> What is the **causal** effect of sector on achievement?

Two interpretations:

--

- Predicting an intervention

    * E.g., what would student $i$'s achievement be if they move to a different type of school?

--

- Counterfactual

    * E.g., what would student $i$'s achievement have been if they had attended a different type of school?

---

# Causal Inference Frameworks

### Potential Outcome Framework (Holland, 1986<sup>1</sup>; Rubin, 1974<sup>2</sup>)

* $Y_{ij}(1) - Y_{ij}(0)$

### Structural Causal Model (Pearl, 2000; 2009<sup>3</sup>)

* $Y_{ij} \mid \mathrm{do}(X = 1) - Y_{ij} \mid \mathrm{do}(X = 0)$

.footnote[

[1] https://doi.org/10.1080/01621459.1986.10478354

[2] https://doi.org/10.1037/h0037350

[3] Pearl, J. (2009). Causality (2nd ed.). 

]

---

# Fundamental Problem of Causal Inference

```{r}
hsb_po |>
    mutate(`mathach (sector = 0)` = if_else(sector == 0, mathach, NA_real_),
           `mathach (sector = 1)` = if_else(sector == 1, mathach, NA_real_)) |>
    select(id, minority, female, ses, sector,
           `mathach (sector = 0)`, `mathach (sector = 1)`) |>
    knitr::kable()
```

---

## Maybe `sector` makes no difference . . . 

```{r}
hsb_po |>
    mutate(`mathach (sector = 0)` = mathach,
           `mathach (sector = 1)` = mathach) |>
    select(id, minority, female, ses, sector,
           `mathach (sector = 0)`, `mathach (sector = 1)`) |>
    mutate(
        `causal effect` = `mathach (sector = 1)` - `mathach (sector = 0)`) |>
    knitr::kable()
```

---

# Confounding

A confounder U is depicted in the following *directed acyclic graph (DAG)*

```{r, out.width = "50%", fig.width = 5}
library(dagitty)
library(ggdag)
dag2 <- dagitty(
    "dag{
      X -> Y; U -> X; U -> Y
      U [unobserved]
    }"
)
coordinates(dag2) <- list(x = c(X = 0, U = 1, Y = 2),
                          y = c(X = 0, U = 1, Y = 0))
# Plot
ggdag(dag2) + theme_dag()
```

U biases the observed association between X and Y from the causal effect of X $\rightarrow$ Y

---

E.g., consider `minority` and `ses` as potential confounders

.pull-left[

### Proportion minority across sectors

```{r}
hsball |>
    group_by(sector) |>
    summarise(minority = mean(minority == 1)) |>
    knitr::kable()
```

]

.pull-right[

### Distribution of `ses` across sectors

```{r, out.width = "95%", fig.width = 6 * .95 / .7 / 2}
library(ggplot2)
ggplot(hsball, aes(x = ses)) +
    geom_histogram() +
    facet_wrap(~ sector, ncol = 1)
```

]

---

# Obtaining Causal Effects

### Randomization

### Unconfounding

---
class: inverse, middle, center

# Randomized Experiments

---

# Why (and When) Does Randomized Experiment Work?

Remove all confounds (probabilistically)

- Intervention groups are different only by chance

```{r, out.width = "60%", fig.width = 6 * .6 / .7}
dag3 <- dagitty(
    "dag{
      X -> Y; U1 -> Y; U2 -> Y
      U1 [unobserved]
      U2 [unobserved]
    }"
)
coordinates(dag3) <- list(x = c(X = 0, U1 = 0.8, U2 = 1.2, Y = 2),
                          y = c(X = 0, U1 = 1, U2 = 1, Y = 0))
# Plot
ggdag(dag3) + theme_dag()
```

---

### Average Treatment Effects

We still don't know the counterfactuals, but
- the distribution of $Y(0)$ should be the same across the "intervention" groups (same for $Y(1)$)

```{r}
hsb_po |>
    mutate(`mathach (sector = 0)` = if_else(sector == 0, mathach, NA_real_),
           `mathach (sector = 1)` = if_else(sector == 1, mathach, NA_real_)) |>
    select(id, minority, female, ses, sector,
           `mathach (sector = 0)`, `mathach (sector = 1)`) |>
    knitr::kable()
```

---
class: inverse, middle, center

# Unconfounding: Statistical Adjustment

---

# Why Do We Include Covariates?

> Statistical control requires causal justification (Wysocki et al., 2022)<sup>1</sup>

.pull-left[

One should adjust for

- Confounders
- Variables blocking confounding paths

]

.pull-right[

```{r, out.width = "100%"}
dag4 <- dagitty(
    "dag{
      X -> Y; W1 -> X; U -> W2; W2 -> X; W1 -> Y; U -> Y
    }"
)
latents(dag4) <- "U"
coordinates(dag4) <- list(x = c(X = 0, W1 = 0.66, U = 1.32, W2 = 0.66, Y = 2),
                          y = c(X = 0, W1 = 1, U = 1, W2 = 0.5, Y = 0))
ggdag(dag4) + theme_dag()
```

]

.footnote[

[1] https://doi.org/10.1177/25152459221095823

]

---

# Statistical Adjustment/Control

.pull-left[

```{r, out.width = "100%", fig.width = 6 * .5 / .7}
ggplot(hsball, aes(x = sector, y = mathach)) +
    geom_boxplot()
```

]

.pull-right[

```{r, out.width = "100%", fig.width = 6 * .5 / .7}
ggplot(hsball, aes(x = factor(ntile(ses, 4)), y = mathach)) +
    geom_boxplot(aes(fill = sector)) +
    labs(x = "SES (quartiles)")
```

]

---

# Causal Inference With Observational Data

- When **all** confounding paths between $X$ and $Y$ are successfully adjusted

- Depends on **causal assumptions**

- When some confounders are not measured, estimated effects are biased

- When wrong variables are adjusted, estimated effects are biased

---

# Confounder vs. Mediator

Do not blindly adjust/control for any variable!

.pull-left[

- Mediator

```{r, out.width = "100%", fig.width = 5}
dag5 <- dagitty(
    "dag{
      X -> Y; X -> M; M -> Y
    }"
)
coordinates(dag5) <- list(x = c(X = 0, M = 1, Y = 2),
                          y = c(X = 0, M = 1, Y = 0))
# Plot
ggdag(dag5) + theme_dag()
```

]

.pull-right[

### Example

Vaccine $\rightarrow$ Antibody $\rightarrow$ Symptom Severity

If adjust for Antibody, may falsely conclude vaccine has no effect

]

---

# So, What to Adjust?

General rule of thumb: if interested in the total effect of $X$, do not adjust for variables that are potential consequences of $X$

Draw a DAG to identify variables on the confounding path

- Preferably, you have identified such variables in the planning stage, so that you can collect data on them

---
class: inverse, middle, center

# Using Multilevel Models for Causal Inference

---

# Student Admissions at UC Berkeley (1973)

.font70[

```{r}
library(tidyr)
berkeley_admit <- UCBAdmissions %>%
  as.data.frame() %>%
  group_by(Gender, Dept) %>%
  mutate(App = sum(Freq)) %>%
  filter(Admit == "Admitted") %>%
  select(Gender, Dept, Admit = Freq, App)
tab <- UCBAdmissions %>%
  as.data.frame() %>%
  group_by(Gender, Dept) %>%
  mutate(App = sum(Freq)) %>%
  filter(Admit == "Admitted") %>%
  ungroup() %>%
  select(Gender, Dept, Admit = Freq, App) %>%
  pivot_wider(names_from = Gender,
              values_from = c(App, Admit))
total_row <- data.frame(
  Dept = "Total",
  App_Male = sum(tab$App_Male),
  Admit_Male = sum(tab$Admit_Male),
  App_Female = sum(tab$App_Female),
  Admit_Female = sum(tab$Admit_Female)
)
bind_rows(tab, total_row) %>%
  mutate(Percent_Male = Admit_Male / App_Male * 100,
         Percent_Female = Admit_Female / App_Female * 100) %>%
  select(1, 2, 4, 6, 3, 5, 7) %>%
  knitr::kable()
```

]

---

# Without Adjustment

```{r, echo = TRUE, output.lines = 10:15}
m1 <- glm(cbind(Admit, App - Admit) ~ Gender,
  data = berkeley_admit,
  family = binomial("logit")
)
summary(m1)
```

---

```{r}
dag8 <- dagitty("dag{
  G -> A; G -> D; D -> A; U -> D; U -> A
}")
coordinates(dag8) <- list(x = c(G = 0, D = 1, A = 2, U = 2),
                          y = c(G = 0, D = 1, A = 0, U = 1))
# Plot
ggdag(dag8) + theme_dag()
```

---

```{r, echo = TRUE, output.lines = 14:24}
berkeley_admit <- berkeley_admit |>
    group_by(Dept) |>
    mutate(Gender_cm = App[2] / sum(App))
m2 <- glmer(cbind(Admit, App - Admit) ~ 
                Gender + Gender_cm + (Gender | Dept),
  data = berkeley_admit,
  family = binomial("logit")
)
summary(m2)
```

---

# The Role of Cluster Means

For level-1 $X$,

including cluster means of $X$ adjusts for differences in $X$ due to cluster-level confounders

```{r, out.width = "60%", fig.width = 6 * .6 / .7}
dag6 <- dagitty(
    "dag{
      U1 -> Xcm; U1 -> Y
      U2 -> X; U2 -> Y
      Xcm -> X; Xcm -> Y
      X -> Y
    }"
)
coordinates(dag6) <- list(x = c(X = 0, Xcm = 0, U1 = 1, U2 = 1, Y = 2),
                          y = c(X = 0, Xcm = 0.8, U1 = 1, U2 = 0.3, Y = 0))
# Plot
ggdag(dag6) + theme_dag()
```

---

# Some Other Useful Tools

- Mediation analysis
    * Whether $X$ has an effect on $Y$ through $M$
    * Check out the `mediation` package

- Propensity score
    * Efficiently balancing multiple covariates

- Instrumental variables (IVs)
    * Variables inducing change in $X$, but should otherwise have no effects on $Y$
    * E.g., the `plm` package can perform IV estimation using the so-called Hausman-Taylor estimator
    
- Causal discovery tools
    * E.g., `pcalg` package