---
title: "HW 9 Template"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
comma <- function(x, d = 2) format(x, digits = d, big.mark = ",")
print_p <- function(x, d = 3) {
    if (x >= .001) {
        return(
            paste("*p* =", sub("^0+", "", format(round(x, d), nsmall = d)))
        )
    } else {
        return("*p* < .001")
    }
}
```

```{r load-pkg, message = FALSE}
library(here)  # makes reading data more consistent
library(tidyverse)  # for data manipulation and plotting
library(haven)  # for importing SPSS/SAS/Stata data
library(lme4)  # for multilevel analysis
library(boot)  # for bootstrap CI
library(bootmlm)  # for multilevel bootstrap
library(performance)  # for ICC and R^2
library(sjPlot)  # for plotting
library(parameters)  # for kr results
library(modelsummary)  # for making tables
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

# Revised Prospectus

You can separate your prospectus and preliminary analysis in different files, if needed. Please Include:

- Sufficient background to understand your research
- Research questions clearly defined
- A description of the nesting structure of your data
- Your plan of data analysis

# Preliminary Analysis

## Import Data

```{r}
happy_dat <- read_sav(here("data_files", "happy_combined.sav"))
# Cluster means
happy_dat <- happy_dat %>%
    group_by(country) %>%
    mutate(income_cm = mean(income)) %>%
    ungroup()
```

## Variable Summary

- `CountryID`: Country ID
- `country`: Country's name
- `income`: Income level (0-_least income_ to 9-_most income_)
- `happy`: Feel happy (1-_not happy_ to 4-_very happy_)

```{r}
# By Country
datasummary(country ~ (income + happy) * (N + Mean + SD),
            data = happy_dat)
```

## Intraclass Correlation

```{r}
m0 <- lmer(happy ~ (1 | country), data = happy_dat)
performance::icc(m0)
```

## Model

Level 1:

$$\text{happy}_{ij} = \beta_{0j} + \beta_{1j} \text{income}_{ij} + e_{ij}$$

Level 2:

$$
  \begin{aligned}
    \beta_{0j} & = \gamma_{00} + \gamma_{01} \text{income\_cm}_{j} + u_{0j}     \\
    \beta_{1j} & = \gamma_{10} + u_{1j}
  \end{aligned}
$$

```{r}
m1 <- lmer(happy ~ income + income_cm + (income | country),
           data = happy_dat)
```

## Results

```{r table-m1}
msummary(m1,
         estimate = c("{estimate} [{conf.low}, {conf.high}]"),
         statistic = NULL,  # suppress the extra rows for SEs
         shape = effect + term ~ model,
         title = "Table 1: Model coefficients")
```

```{r fig-m1}
plot_model(m1, type = "pred", pred.type = "re",
           terms = "income", show.data = TRUE,
           jitter = 0.1, dot.size = 0.2)
```

```{r, include = FALSE}
m1_coef <- parameters::parameters(m1, ci_method = "satterthwaite")
print_coef <- function(term, tab = m1_coef) {
    coef_pos <- which(tab$Parameter == term)
    paste0(comma(tab[coef_pos, "Coefficient"]),
           ", *SE* = ", comma(tab[coef_pos, "SE"]),
           ", *t*(", comma(tab[coef_pos, "df_error"]), ") = ",
           comma(tab[coef_pos, "t"], 3),
           ", ", print_p(tab[coef_pos, "p"]))
}
r2_m1 <- MuMIn::r.squaredGLMM(m1)
```

```{r, include = FALSE, cache = TRUE}
boo_r2 <- bootstrap_mer(m1, MuMIn::r.squaredGLMM, nsim = 1999,
                        type = "residual")
```

Table 1 shows the fixed-effect coefficients. We found evidence for the positive association, averaged across countries, between income and happiness at the individual level, $\gamma_{10}$ = `r print_coef("income")`. The contextual effect was not significant, $\gamma_{01}$ = `r print_coef("income_cm")` (see the Figure). The marginal $R^2$ (Nakagawa et al., 2017) for the model was estimated to be `r comma(r2_m1[1], 2)`, 95% bootstrap CI [`r paste0(round(boot.ci(boo_r2, index = 1, type = "perc")$percent[4:5], 2), collapse = ", ")`].